---
title: "Whole Game: Data Transformation"
subtitle: "R for Data Science (2e)"
author: "Clare Gibson"
date: "23-10-2024"
output:
  html_document:
    theme: yeti
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
# Knitr chunk options
knitr::opts_chunk$set(
	echo = TRUE,
	fig.align = "center",
	message = FALSE,
	warning = FALSE
)
```

# 3.1 Introduction

The goal of this chapter is to give you an overview of all the key tools for transforming a data frame.

## Prerequisites

Load the required packages.

```{r load-packages}
library(tidyverse)
library(nycflights13)
```

## nycflights13

Take a look at the data

```{r view-data}
flights
```

## dplyr basics

You’re about to learn the primary dplyr verbs (functions), which will allow you to solve the vast majority of your data manipulation challenges. But before we discuss their individual differences, it’s worth stating what they have in common:

1.  The first argument is always a data frame.
2.  The subsequent arguments typically describe which columns to operate on using the variable names (without quotes).
3.  The output is always a new data frame.

# 3.2 Rows

The most important verbs that operate on rows of a dataset are `filter()`, which changes which rows are present without changing their order, and `arrange()`, which changes the order of the rows without changing which are present. Both functions only affect the rows, and the columns are left unchanged. We’ll also discuss `distinct()` which finds rows with unique values. Unlike `arrange()` and `filter()` it can also optionally modify the columns.

## Exercises
In a single pipeline for each condition, find all flights that meet the condition:

- Had an arrival delay of two or more hours
```{r ex-3-2-1-1}
flights |> 
  filter(arr_delay >= 120)
```

- Flew to Houston (IAH or HOU)
```{r ex-3-2-1-2}
flights |> 
  filter(dest %in% c("IAH", "HOU"))
```

- Were operated by United, American, or Delta
```{r ex-3-2-1-3}
flights |> 
  filter(carrier %in% c("UA", "AA", "DL"))
```

- Departed in summer (July, August, and September)
```{r ex-3-2-1-4}
flights |> 
  filter(month %in% c(7,8,9))
```

- Arrived more than two hours late but didn’t leave late
```{r ex-3-2-1-5}
flights |> 
  filter(
    arr_delay > 120,
    dep_delay <= 0
  )
```

- Were delayed by at least an hour, but made up over 30 minutes in flight
```{r ex-3-2-1-6}
flights |> 
  filter(
    dep_delay >= 60,
    arr_delay < 30
  )
```

2. Sort `flights` to find the flights with the longest departure delays. Find the flights that left earliest in the morning.
```{r ex-3-2-2}
flights |> 
  arrange(desc(dep_delay))

flights |> 
  arrange(dep_time)
```

3. Sort `flights` to find the fastest flights. (Hint: Try including a math calculation inside of your function.)
```{r ex-3-2-3}
flights |> 
  arrange(distance/air_time)
```

4. Was there a flight on every day of 2013?
```{r ex-3-2-4}
flights |> 
  filter(year == 2013) |> 
  distinct(year, month, day)
```

5. Which flights traveled the farthest distance? Which traveled the least distance?
```{r ex-3-2-5}
flights |> 
  arrange(desc(distance)) |> 
  head(10)
```

6. Does it matter what order you used `filter()` and `arrange()` if you’re using both? Why/why not? Think about the results and how much work the functions would have to do.
```{r ex-3-2-6}
flights |> 
  filter(month == 7) |> 
  arrange(distance, year, month, day)
```

Filtering first will lead to faster calculation but the order does not change the result.

# 3.3 Columns

There are four important verbs that affect the columns without changing the rows: `mutate()` creates new columns that are derived from the existing columns, `select()` changes which columns are present, `rename()` changes the names of the columns, and `relocate()` changes the positions of the columns.

## Exercises

1. Compare `dep_time`, `sched_dep_time`, and `dep_delay`. How would you expect those three numbers to be related?
```{r ex-3-3-1}
flights |> 
  select(
    dep_time,
    sched_dep_time,
    dep_delay
  )
```

As expected, `dep_delay` is the result of subtracting `dep_time` from `sched_dep_time`.

2. Brainstorm as many ways as possible to select `dep_time`, `dep_delay`, `arr_time`, and `arr_delay` from flights.
```{r ex-3-3-2}
flights |> 
  select(starts_with("dep_") | starts_with("arr_"))
```

3. What happens if you specify the name of the same variable multiple times in a `select()` call?
```{r ex-3-3-3}
flights |> 
  select(dep_delay, dep_delay)
```

R only returns the column once.

4. What does the `any_of()` function do? Why might it be helpful in conjunction with this vector?
```{r ex-3-3-4}
variables <- c("year", "month", "day", "dep_delay", "arr_delay")
flights |> 
  select(any_of(variables))
```

5. Does the result of running the following code surprise you? How do the select helpers deal with upper and lower case by default? How can you change that default?
```{r ex-3-3-5}
flights |> select(contains("TIME"))
```

By default, the `ignore.case` argument is set to TRUE, so case is ignored. You can change that behavious by setting the argument to FALSE.

6. Rename `air_time` to `air_time_min` to indicate units of measurement and move it to the beginning of the data frame.
```{r ex-3-3-6}
flights |> 
  rename(air_time_min = air_time) |> 
  relocate(air_time_min)
```

7. Why doesn’t the following work, and what does the error mean?
```{r ex-3-3-7}
flights |> 
  select(tailnum, arr_delay) |> 
  arrange(arr_delay)
```

You can't arrange by a column that isn't included in the `select` function.

# 3.5 Groups
```{r group-by}
flights |> 
  group_by(year, month, day) |> 
  summarise(avg_arr_delay = mean(arr_delay, na.rm = TRUE), .groups = "drop")
```

## Exercises
1. Which carrier has the worst average delays? Challenge: can you disentangle the effects of bad airports vs. bad carriers? Why/why not? (Hint: think about `flights |> group_by(carrier, dest) |> summarize(n())`).
```{r ex-3-5-1-1}
flights |> 
  group_by(carrier) |> 
  summarise(avg_arr_delay = mean(arr_delay, na.rm = TRUE)) |> 
  arrange(desc(avg_arr_delay))
```

```{r ex-3-5-1-2}
flights |> 
  group_by(carrier, dest) |> 
  summarise(n = n())
```

2. Find the flights that are most delayed upon departure from each destination.
```{r ex-3-5-2}
flights |> 
  slice_max(dep_delay, by = dest, na_rm = TRUE) |> 
  arrange(dest) |> 
  select(dest, carrier, flight, tailnum, origin, dest, dep_delay)
```

3. How do delays vary over the course of the day? Illustrate your answer with a plot.
```{r ex-3-5-3}
avg_dep_delay_by_hour <- 
  flights |> 
  group_by(hour) |> 
  summarise(avg_dep_delay = mean(dep_delay, na.rm = TRUE))

avg_dep_delay_by_hour |> 
  ggplot(
    aes(
     x = hour,
     y = avg_dep_delay
    )
  ) +
  geom_smooth()
```

During the course of a day, the average departure delay increases until around 7pm then declines until the end of the day.

4. What happens if you supply a negative n to `slice_min()` and friends?
```{r ex-3-5-4}
flights |> 
  slice_min(dep_delay, n = -5) |> 
  select(hour, dep_delay)
```

Supplying a negative number simply arranges the tibble in descending order without filtering the rows.

5. Explain what `count()` does in terms of the dplyr verbs you just learned. What does the `sort` argument to `count()` do?
```{r ex-3-5-5}
flights |> 
  count(origin, sort = TRUE)
```

Using the `count()` function is the equivalent of grouping and then counting each group. The `sort` argument arranges the output in descending order of count.

6. Suppose we have the following tiny data frame:
```{r ex-3-5-6}
df <- tibble(
  x = 1:5,
  y = c("a", "b", "a", "a", "b"),
  z = c("K", "K", "L", "L", "K")
)
```

- Write down what you think the output will look like, then check if you were correct, and describe what group_by() does.
```{r ex-3-5-6-1}
df |>
  group_by(y)
```

This code will return an unaltered tibble but behind the scenes R will be ready to perform the next operation on groups according to `y`.

- Write down what you think the output will look like, then check if you were correct, and describe what `arrange()` does. Also, comment on how it’s different from the `group_by()` in part (a).
```{r ex-3-5-6-2}
df |>
  arrange(y)
```

This will rearrange the rows in `df` in ascending order of `y`.

- Write down what you think the output will look like, then check if you were correct, and describe what the pipeline does.
```{r ex-3-5-6-3}
df |>
  group_by(y) |>
  summarize(mean_x = mean(x))
```

For every group of `y` we will see the mean of `x`.

- Write down what you think the output will look like, then check if you were correct, and describe what the pipeline does. Then, comment on what the message says.
```{r ex-3-5-6-4}
df |>
  group_by(y, z) |>
  summarize(mean_x = mean(x))
```

It will group the output by `y` then `z` and for each group it will give the mean of `x`.

- Write down what you think the output will look like, then check if you were correct, and describe what the pipeline does. How is the output different from the one in part (d)?
```{r ex-3-5-6-5}
df |>
  group_by(y, z) |>
  summarize(mean_x = mean(x), .groups = "drop")
```

This does the same as above but the groups are dropped afterwards.

- Write down what you think the outputs will look like, then check if you were correct, and describe what each pipeline does. How are the outputs of the two pipelines different?
```{r ex-3-5-6-6}
df |>
  group_by(y, z) |>
  summarize(mean_x = mean(x))

df |>
  group_by(y, z) |>
  mutate(mean_x = mean(x))
```

The first output will include one row per group. The second will include one row per observation but the aggregated measure will be duplicated for each row in the group.